---
title: "NYPD Shooting Final"
author: "Julio Tapia"
date: "2025-02-28"
output:
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)
library(tidyverse)
library(lubridate)
library(ggplot2)
```

## Importing the Data

```{r import_data, message=FALSE}
# URL for the dataset
nypd_data_url <- "https://data.cityofnewyork.us/api/views/833y-fsy8/rows.csv?accessType=DOWNLOAD"

# Load the dataset
df <- read_csv(nypd_data_url)
```

## Data Preprocessing

```{r clean_data}
# Convert date column to proper format
df <- df %>%
  mutate(OCCUR_DATE = mdy(OCCUR_DATE),
         Year = year(OCCUR_DATE),
         Month = month(OCCUR_DATE, label = FALSE)) # Ensure Month is numeric

# Standardizing victim age groups
df <- df %>%
  mutate(VIC_AGE_GROUP = recode(VIC_AGE_GROUP,
                                "<18" = "Under 18",
                                "18-24" = "18-24",
                                "25-44" = "25-44",
                                "45-64" = "45-64",
                                "65+" = "65+",
                                "UNKNOWN" = "Unknown"))
```

## Visualization 1: Shooting Incidents Over Time

```{r lineplot_shootings, echo=FALSE}
shooting_trend <- df %>%
  count(Year, Month) %>%
  arrange(Year, Month)

ggplot(shooting_trend, aes(x = as.Date(paste(Year, Month, "01", sep = "-")), y = n)) +
  geom_line(color = "blue") +
  labs(title = "NYPD Shooting Incidents Over Time",
       x = "Year",
       y = "Number of Incidents") +
  theme_minimal()
```

## Visualization 2: Shootings by Borough

```{r barplot_borough, echo=FALSE}
borough_count <- df %>% count(BORO) %>% filter(!is.na(BORO))

ggplot(borough_count, aes(x = reorder(BORO, -n), y = n, fill = BORO)) +
  geom_bar(stat = "identity") +
  labs(title = "Number of Shootings by Borough",
       x = "Borough",
       y = "Number of Incidents") +
  theme_minimal() +
  scale_fill_brewer(palette = "Set2")
```

## Visualization 3: Victim Demographics (Age Group)

```{r barplot_victim_age, echo=FALSE}
victim_age_dist <- df %>% count(VIC_AGE_GROUP) %>% filter(!is.na(VIC_AGE_GROUP))

ggplot(victim_age_dist, aes(x = reorder(VIC_AGE_GROUP, -n), y = n, fill = VIC_AGE_GROUP)) +
  geom_bar(stat = "identity") +
  labs(title = "Victim Age Group Distribution",
       x = "Age Group",
       y = "Number of Victims") +
  theme_minimal() +
  scale_fill_brewer(palette = "Set1") +
  coord_flip()
```

## Visualization 4: Shootings by Month

```{r barplot_monthly_trends, echo=FALSE}
monthly_trend <- df %>% count(Month) %>% arrange(Month)

ggplot(monthly_trend, aes(x = factor(Month, levels = 1:12, labels = month.abb), y = n, fill = factor(Month))) +
  geom_bar(stat = "identity") +
  labs(title = "NYPD Shooting Incidents by Month",
       x = "Month",
       y = "Number of Incidents") +
  theme_minimal() +
  scale_fill_brewer(palette = "Blues")
```

## Conclusion

- **Trends Over Time:** Incidents fluctuate yearly, with noticeable spikes in certain periods.
- **Borough Impact:** Brooklyn and the Bronx see the highest incidents, while Staten Island has the lowest.
- **Victim Demographics:** Young adults (18-44) are most affected.
- **Seasonal Trends:** Shootings peak in summer (June-August) and drop in winter.

**This analysis demonstrates how data science can uncover crime patterns, helping policymakers allocate resources effectively.**
